---
title: "Lab D"
author: "Mujie Wang"
date: "11/22/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(survival)
library(MASS)
hern1 <- read.table("/Users/wangmujie/Desktop/hern.txt",header = TRUE)
retire <- read.table("/Users/wangmujie/Desktop/retire.txt", header = TRUE)
```


1(a)
```{r}
hern1.surv <- Surv(hern1$Time, hern1$Status)
hern1.pr <- coxph(hern1.surv ~ hern1$Prison)
summary(hern1.pr)
```
The p-value of prison is 0.263 which is greater than alpha = 0.05,
there is no significant effect.

1(b)
```{r}
hern1.pr.cl <- coxph(hern1.surv ~ Clinic+Prison, data = hern1) 
hern1.cl <- coxph(hern1.surv ~ Clinic, data = hern1) 
lrt.prison <- 2*(hern1.pr.cl$loglik[2]-hern1.cl$loglik[2]) 
pchisq(lrt.prison, df= 1, lower.tail = FALSE)
```
Since the p-value is 0.09493 which is gereater than a = 0.05, which means that the prison
time is not a significant given the clinic is included in the model.


1(c)
```{r}
hern1.fit.cl <- survfit(hern1.surv ~ Clinic, data = hern1)

plot(hern1.fit.cl,fun="cloglog", main="Log-log plot of Heroin Addicts from Two Clinics",
     xlab = "days in treatment program", ylab = "log(-log S(t))",
     lwd=2, col=2:1, mark.time=TRUE, mark = 20)

legend("topleft",inset = 0.10,legend=c("Clinic 1", "Clinic2"),
       pch=15,col=2:1)
```
In the plot there two line represent the clinics, and are two crossing point. At the same
time, the lines are diverged as the time of the treatment program increases.

1(d)
```{r}
hern1.pr.stcl <- coxph(hern1.surv ~ Prison+ strata(Clinic), data = hern1)
summary(hern1.pr.stcl)
```
The p-value is 0.045, which is less than a = 0.05. It mean that the prison time have
effect. In part 1b, we assime that the baseline hazard rate is the same for all clinics,
but this model we use strata(Clinic), so it make the constant hazard ratio more
significant with Prison.


1(e)
```{r}
hern1.pr_stCl <- coxph(hern1.surv ~ Prison*strata(Clinic), data = hern1)
summary(hern1.pr_stCl)

lrt.prC <- 2*(hern1.pr_stCl$loglik[2]-hern1.pr.stcl$loglik[2]) 
pchisq(lrt.prC, df=1,lower.tail = FALSE)

anova(hern1.pr_stCl,hern1.pr.stcl)
```
This is a test about significant interaction between the prison variable and the clinic
variable, and the p-value is 0.3966961 which is greater than 0.05, means that they are
not significant. The interaction term means that the subject the prison time or not and
which clinic they go to, and see how long they stays in the treatment program for the
heroin addiction. It can also help use to decide if one variable depended on the other.


2(a)
```{r}
retire.surv <- Surv(retire$time, retire$death) 
retire.ge <- coxph(retire.surv ~ gender, data = retire) 
summary(retire.ge)
```
The likelihood ratio statistic is 5.78 and the p-value is 0.016 which is less then 0.05, 
so that the survival time between 2 genders has a significant different.

2(b)
```{r}
retire.a <- coxph(retire.surv ~ ageentry, data = retire)
retire.g.a <- coxph(retire.surv ~ gender + ageentry, data = retire) 
anova(retire.a,retire.g.a )
```
The p-value is 0.03406 which is less then 0.05, so there is a significant difference 
in the survival time between different gender (men and women). 
However, compare to part a, they have a different model fitting. 
In part a, it did not control the ageentry;
In part b, it did the age will keep constant after they entered the facility.

2(c)
```{r}
retire.g_a <- coxph(retire.surv ~ gender + age + gender*age, data = retire)
lrt.retire <- 2*(retire.g_a$loglik[2] - retire.ge$loglik[2]) 
pchisq(lrt.retire, df=1, lower.tail = FALSE)

retire.g_ag <- coxph(retire.surv ~ gender + ageentry + gender*ageentry, data = retire)
lrt.retire_ag <- 2*(retire.g_ag$loglik[2] - retire.ge$loglik[2]) 
pchisq(lrt.retire_ag, df=1, lower.tail = FALSE)
```
For age: The interaction between age and gender is signifocant or not 
is showed why using the likelihood ratio test and the interaction is significant 
because the p-value is 0.0014 which is less than 0.05.
For ageentry: The interaction between age and gender is signifocant or not 
is showed why using the likelihood ratio test and the interaction is significant 
because the p-value is 2.137116e-11 which is less than 0.05.

2(d)
```{r}
retire.fit <- survfit(retire.surv ~ gender, data = retire) 
plot(retire.fit,fun="cloglog", main="Senior Care Facility for Men and Women",
     xlab = "months spent at the facility", ylab = "log(-log S(t))", lwd=2,
     col=2:1, mark.time=TRUE, mark = 20)
legend("topleft",inset = 0.15,legend=c("Men", "Women"),pch=15,col=2:1)
```
The crossing lines show that the proportional hazards assumption is not reasonable for the model. 
This plot does not satisfy the constant distance between the differentce 
in the log(-log(s(t))) that we want so have.

2(e)
We use ageentry as our covariate and not age which is the age of the subject when the event occurred
because we don???t want to assume that younger individuals have a longer survival time than the older. 
It is because there factors are affecting the survival time such as illness they have before. 
We also want to focus on how the time of individual stay in the senior care affect their survivial time.













3(a)
```{r}
retire_3 <- retire
retire_3$time[retire_3$time == 0] = 0.05
retire_3.split <- survSplit(Surv(time,death)~., episode = "episode",
                           cut = 48, id = "sub.id", data = retire_3)
retire_3.split[1:5, ]
```

3(b)
```{r}
retire_3.splitfit <- coxph(Surv(tstart, time, death)~ gender:strata(episode),
                            data = retire_3.split)
retire_3.splitfit2 <- coxph(Surv(tstart, time, death)~ ageentry + gender:strata(episode),
                            data = retire_3.split)
lrt.retire_3split <- 2*(retire_3.splitfit2$loglik[2]-retire_3.splitfit$loglik[2])
pchisq(lrt.retire_3split, df=1, lower.tail = FALSE)
```

```{r}
summary(retire_3.splitfit2)
anova(retire_3.splitfit,retire_3.splitfit2)
```
3(c)
```{r}
ep1 <- subset(retire_3.split, retire_3.split$episode==1) 
ep2 <- subset(retire_3.split, retire_3.split$episode==2)

ep1.coxph <- coxph(Surv(time,death)~gender, data = ep1) 
summary(ep1.coxph)

ep2.coxph <- coxph(Surv(time,death)~gender, data = ep2) 
summary(ep2.coxph)
```
The 95% CI of the hazard ratio for men before 48 months cutoff is [0.3092, 0.8785] and after 48 months
cutoff is [0.4851, 1.181].

3(d)
I conclude that it is important to consider a change in the effect of gender before and after 4 years in the
retirement facility, in the 95% CI for the hazard ratio before the 4 years cutoff from part 3c the result is
[0.3092, 0.8785], so we can conclude what the effect of gender is siginificat to the model. While the after
4 years cut off, the 95% CI is [0.4851, 1.181], which show that the effect of gender is not significant
because 1 is inluded. I think that it maybe because people in different gender will have similar treatment
so there are less different betern the hazard rate and survival time for different gender.